Missing the Point\(\colon\) Non-Convergence in Multiple Imputation Procedures

### 2020-04-02 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Put different percentages missingness in the manuscript

---
title: Missing the Point$\colon$ Non-Convergence in Multiple Imputation Procedures
runninghead: Oberman
author:
- name: H. I. Oberman
  num: 1
address:
- num: 1
  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands
corrauth: "Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands."
email: h.i.oberman@uu.nl
abstract: "Multple imputation by chained equations (mice) is a widely used tool to accommodate missing data. While empirical evidence supports the validity of inferences obtained using mice, there is no consensus on the convergence properties of the method. This paper provides insight into non-convergence of mice algorithms."
keywords: MICE; convergence 
classoption:
  - Royal
  - times
bibliography: thesis.bib
bibliographystyle: sageh
output:
  rticles::sage_article:
    keep_tex: yes
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 5, echo = FALSE, fig.align = "center") 
load("../Results/results.Rdata")
source("../2.CreateFigures.R")

@@

At some point, any scientist conducting statistical analyses will run into a missing data problem [@alli02]. Missingness is problematic because statistical inference cannot be performed on incomplete data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which may yield wildly invalid results [@buur18]. A popular answer to the ubiquitous problem of missing information is to use the framework of multiple imputation (MI), proposed by @rubin87. **(Call it mice already, then explain mi and ce sides of mice)** For the theoretical foundation of MI, see @rubin87. For an accessible and comprehensive introduction to MI from an applied perspective, see e.g. @buur18. With MI, missing datapoints are 'imputed' (i.e., filled in) via an *imputation model* (**term not required to understand this paper, so circumvent**) . (**Move this to mi vs. ce part in intro:** The imputation model can be a joint model for all variables, or several univariate conditional distributions (fully conditional specification). FCS is more flexible and "avoids the problem of specifying an appropriate joint imputation distribution and replaces this by the selection of appropriate univariate conditional distributions".)**

**(Rewrite: With FCS, an iterative algorithmic procedure is used to ... multiple imputation by chained equations ... "The values for imputing the missing data are drawn from the posterior predictive distribution of the missing data conditional on the observed data.")**

@@

Current practice is visually inspecting imputation chains for signs of non-convergence, but may be challenging to the untrained eye [@buur18, $\S$ 6.5.2]. Therefore diagnostic evaluation of convergence would be preferred. However, not known whether MCMC diagnostics are appropriate for MICE ... The application of convergence diagnostics to MI methods has not been systematically studied [@buur18]. How can non-convergence be diagnosed? How many iterations are sufficient/required? What are the effects of non-convergence on estimates, predictions and inference? What is an appropriate non-convergence diagnostic? Are the default number of iterations sufficient? Defaults in: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100?? [(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 10 [(link)](https://www.stata.com/manuals13/mi.pdf, p. 139), Amelia = NA, because resampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf).

@@

Let $y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, \dots, n$) on variable $j$ ($j = 1, \dots, p$) may be either observed or missing. The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$. **(This is not only notation, but theory too: For each datapoint in $y_{mis}$ $M \times T$ times plausible values are sampled, where $M$ is the number of imputations ($m = 1, \dots, M$) and $T$ is the number of iterations ($t = 1, \dots, T$).)** The collection of samples between the initial value (at $t=1$) and the imputed value (at $t=T$) will be referred to as an 'imputation chain'.

@@

@@

This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms. We assess the performance of the two convergence diagnostics against the recommended evaluation criteria for MI methods [i.e., average bias, average confidence interval width, and empirical coverage rate across simulations; @buur18, $\S$ 2.5.2]. That is, there is no baseline measure available to evaluate performance against. **(Move above: The current practice of visually inspecting imputation chains for signs of non-mixing or non-stationarity can only diagnose severely pathological cases of non-convergence.)**
<!-- We could also look at distributional characteristics, and plausibility of imputed values, see @vinknd. For now, this is outside of the scope of this study. -->

@@

The complete data is 'amputed' once for each simulation repetition. That is, the function `mice::ampute()` is used to impose a missingness mechanism upon the data. **(change this: The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`). This leaves 20\% of the rows completely observed). **

@@

**(tot hier doorgelezen)**

@@

where $t_{(m-1)}$ is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate. From bias and CIW, we calculate empirical coverage rates. Coverage rate is the proportion of simulations in which $Q$ is between the bounds of the CI95\% around $\bar{Q}$.

@@

```{r, fig.cap="Convergence diagnostics (means)"}
bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3, guides = "collect")
```

```{r, fig.cap="Convergence diagnostics (variances)"}
bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3, guides = "collect")
```

```{r, fig.cap="Convergence diagnostics (regression coefficients)"}
bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3, guides = "collect")
```

## Convergence Diagnostics (pair convergnce diagnostics with respective estimates, not split between convergence and simulation diagnostics)

**(Add different percentages missingness??)**

@@

According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values. The negative $AC$-values, however, indicate increasing divergence (**move above!**) between subsequent samples in conditions where $T<4$. It may not be wise (**why? and how would we know? e.g., wider distrubutions? didn't seem like it...**) to terminating the algorithm at that point, but after `recovering' from this dip in $AC$-values. This would imply $T=6$ as the minimal number of iterations.

@@

CIW is only clearly divergent in the simulation condition $T=1$, see Figure 3B. However, under-estimating the variance of $\bar{Q}$, which is the case for $T=3$, may yield spurious inferences (**move above**). Only conditions where $T>3$ are therefore considered sufficient. (**Possibly irrelevant Loess: These conditions have similar CIWs, but across iterations stability is not established (i.e., the Loess line is never completely flat within the 100 simulation conditions).**)

The empirical coverage rate across repetitions seems more or less stable for conditions where $T>1$, see Figure 3C. We see some over-coverage at $T=2$, but that is better than under-estimating the variance of $\bar{Q}$. On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% [@neym34] (**not anymore??**). (**Loess: Similar to the CIWs, there is some trending across iterations (i.e., the Loess line is never flat)**).

@@

The observed dip in AC implies that default maxit value of five iterations is the worst possible number of iterations. Moreover, the results of this study imply that assessing the stationarity component of convergence with $AC$ might be redundant, since high $AC$-values are implausible in MI procedures. That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains (**add above!!**). $AC$ would thus not be informative of the convergence of MI algorithms.

(**Add above, not here because it was a known limitation: Since this study only considers only an MCAR missingness mechanism, results may not be extrapolated to other missing data problems. Proper performance of the convergence diagnostics under MCAR is necessary but not sufficient to demonstrate appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics. This is just a proof of concept.**)

>>>>

---
title: Missing the Point$\colon$ Non-Convergence in Multiple Imputation Procedures
runninghead: Oberman
author:
- name: H. I. Oberman
  num: 1
address:
- num: 1
  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands
corrauth: "Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands."
email: h.i.oberman@uu.nl
abstract: "(**Rewrite:**) Multple imputation by chained equations (mice) is a widely used tool to accommodate missing data. While empirical evidence supports the validity of inferences obtained using mice, there is no consensus on the convergence properties of the method. This paper provides insight into non-convergence of mice algorithms."
keywords: MICE; convergence 
classoption:
  - Royal
  - times
bibliography: thesis.bib
bibliographystyle: sageh
output:
  rticles::sage_article:
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 4.8, fig.height = 3.6, echo = FALSE, fig.align = "left") 
```

@@

At some point, any social scientist 
<!-- conducting statistical analyses  -->
will run into a missing data problem [@alli02]. Missingness is problematic because statistical inference cannot be performed on incomplete data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which may yield wildly invalid results [@buur18]. A popular method to obtain statistically valid results despite of missing data is 'multiple imputation' [MI; @rubin87]. This paper investigates a potential threat to the validity of this method: algorithmic non-convergence.

With multiple imputation, missing data are 'imputed' (i.e., filled in) several times by means of an iterative algorithmic procedure. The method ultimately results in an aggregated statistical inference. For the theoretical foundation of MI, see @rubin87. For an accessible and comprehensive introduction to MI from an applied perspective, see e.g. @buur18.

based on a model of the observed databy means of an iterative algorithm 
is the use of MICE: multiple imputation by chained equations. MICE is a type of multiple imputation procedure--a framework proposed by @rubin87--in which missing datapoints are 'imputed' (i.e., filled in) by means of an iterative algorithm (chained equations). Imputed values are obtained by modeling the (distribution of the) missing data conditional on the observed data. The technique is therefore also called 'fully conditional specification' (FCS).
<!-- several times, and inferences are aggregated across imputed datasets.  -->
<!-- and statistical inference is performed on each imputed dataset. Variation between the results refelect uncertainty due to missingness  -->
<!-- CE: The imputation model can be a joint model for all variables, or several univariate conditional distributions (fully conditional specification). FCS is more flexible and "avoids the problem of specifying an appropriate joint imputation distribution and replaces this by the selection of appropriate univariate conditional distributions".  -->

<!-- (**Rewrite:** With FCS, an iterative algorithmic procedure is used to ... multiple imputation by chained equations ... "The values for imputing the missing data are drawn from the posterior predictive distribution of the missing data conditional on the observed data.")  -->

@@

Current practice is visually inspecting imputation chains for signs of non-convergence, but may be challenging to the untrained eye [@buur18, $\S$ 6.5.2]. Visually inspecting imputation chains (**for signs of non-mixing or non-stationarity**) can only diagnose severely pathological cases of non-convergence. Therefore diagnostic evaluation of convergence would be preferred. However, not known whether MCMC diagnostics are appropriate for MICE ... The application of convergence diagnostics to MI methods has not been systematically studied [@buur18]. How can non-convergence be diagnosed? How many iterations are sufficient/required? What are the effects of non-convergence on estimates, predictions and inference? What is an appropriate non-convergence diagnostic? Are the default number of iterations sufficient? Defaults in: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100?? [(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 10 [(link)](https://www.stata.com/manuals13/mi.pdf, p. 139), Amelia = NA, because resampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf).

@@

Let $y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, \dots, n$) on variable $j$ ($j = 1, \dots, p$) may be either observed or missing. The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$. (**This is not only notation, but theory too:** For each datapoint in $y_{mis}$ $M \times T$ times plausible values are sampled, where $M$ is the number of imputations ($m = 1, \dots, M$) and $T$ is the number of iterations ($t = 1, \dots, T$).) The collection of samples between the initial value (at $t=1$) and the imputed value (at $t=T$) will be referred to as an 'imputation chain'.

@@

Negative $AC$-values indicate divergence between subsequent samples. It may not be wise (**why? and how would we know? e.g., wider distrubutions? didn't seem like it...**) to terminating the algorithm when AC is negative. High $AC$-values are implausible in MI procedures. That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.

@@

This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms. We assess the performance of the two convergence diagnostics against the recommended evaluation criteria for MI methods [i.e., average bias, average confidence interval width, and empirical coverage rate across simulations; @buur18, $\S$ 2.5.2]. That is, there is no baseline measure available to evaluate performance against.
<!-- We could also look at distributional characteristics, and plausibility of imputed values, see @vinknd. For now, this is outside of the scope of this study. -->

@@

The complete data is 'amputed' once for each simulation repetition. That is, the function `mice::ampute()` is used to impose a missingness mechanism upon the data. (**change this:** The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`). This leaves 20\% of the rows completely observed). **(Add different percentages missingness??)** This study only considers only an MCAR missingness mechanism. Therefore, results may not be extrapolated to other missing data problems. Proper performance of the convergence diagnostics under MCAR is necessary but not sufficient to demonstrate appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics. This is just a proof of concept.

@@

@@

where $t_{(m-1)}$ is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate. **Why track CIW?** Under-estimating the variance of $\bar{Q}$ may yield spurious inferences.
From bias and CIW, we calculate empirical coverage rates. Coverage rate is the proportion of simulations in which $Q$ is between the bounds of the CI95\% around $\bar{Q}$.

@@

```{r, include = FALSE}
load("../Results/25percentmiss.Rdata")
source("../2.CreateFigures.R")
```

```{r}
bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of means, 25% missingness")
```

```{r}
bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of variances, 25% missingness")
```

```{r}
bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of effects, 25% missingness")

```{r, include = FALSE}
load("../Results/50percentmiss.Rdata")
source("../2.CreateFigures.R")
```

```{r}
bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of means, 50% missingness")
```

```{r}
bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of variances, 50% missingness")
```

```{r}
bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of effects, 50% missingness")

```{r, include = FALSE}
load("../Results/75percentmiss.Rdata")
source("../2.CreateFigures.R")
```

```{r}
bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of means, 75% missingness")
```

```{r}
bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of variances, 75% missingness")
```

```{r}
bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of effects, 75% missingness")

```{r, include = FALSE}
load("../Results/95percentmiss.Rdata")
source("../2.CreateFigures.R")
```

```{r}
bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of means, 95% missingness")
```

```{r}
bias_sds_plot + R_sds_plot + AC_sds_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of variances, 95% missingness")
```

```{r}
bias_ests_plot + bias_R_sq_plot + cov_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of effects, 95% missingness")

## Convergence Diagnostics

**(Pair convergnce diagnostics with respective estimates, not split between convergence and simulation diagnostics)**

@@

According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values.

@@

CIW is only clearly divergent in the simulation condition $T=1$, see Figure 3B. Only conditions where $T>3$ are therefore considered sufficient. (**Possibly irrelevant Loess:** These conditions have similar CIWs, but across iterations stability is not established (i.e., the Loess line is never completely flat within the 100 simulation conditions).)

The empirical coverage rate across repetitions seems more or less stable for conditions where $T>1$, see Figure 3C. We see some over-coverage at $T=2$, but that is better than under-estimating the variance of $\bar{Q}$. On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% [@neym34] (**not anymore??**). (**Loess:** Similar to the CIWs, there is some trending across iterations (i.e., the Loess line is never flat)).

@@

The observed dip in AC implies that default maxit value of five iterations is the worst possible number of iterations. Moreover, the results of this study imply that assessing the stationarity component of convergence with $AC$ might be redundant. 
<!-- $AC$ would thus not be informative of the convergence of MI algorithms.  -->

### 2020-04-02 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Put different percentages missingness in the manuscript

\begin{abstract}
Multple imputation by chained equations (mice) is a widely used tool to
accommodate missing data. While empirical evidence supports the validity
of inferences obtained using mice, there is no consensus on the
convergence properties of the method. This paper provides insight into
non-convergence of mice algorithms.
\end{abstract}

@@

At some point, any scientist conducting statistical analyses will run
into a missing data problem \citep{alli02}. Missingness is problematic
because statistical inference cannot be performed on incomplete data
without employing \emph{ad hoc} solutions (e.g., list-wise deletion),
which may yield wildly invalid results \citep{buur18}. A popular answer
to the ubiquitous problem of missing information is to use the framework
of multiple imputation (MI), proposed by \citet{rubin87}. \textbf{(Call
it mice already, then explain mi and ce sides of mice)} For the
theoretical foundation of MI, see \citet{rubin87}. For an accessible and
comprehensive introduction to MI from an applied perspective, see
e.g.~\citet{buur18}. With MI, missing datapoints are `imputed' (i.e.,
filled in) via an \emph{imputation model} (\textbf{term not required to
understand this paper, so circumvent}) . (\textbf{Move this to mi vs.~ce
part in intro:} The imputation model can be a joint model for all
variables, or several univariate conditional distributions (fully
conditional specification). FCS is more flexible and ``avoids the
problem of specifying an appropriate joint imputation distribution and
replaces this by the selection of appropriate univariate conditional
distributions''.)**

\textbf{(Rewrite: With FCS, an iterative algorithmic procedure is used
to \ldots{} multiple imputation by chained equations \ldots{} ``The
values for imputing the missing data are drawn from the posterior
predictive distribution of the missing data conditional on the observed
data.'')}

@@

Current practice is visually inspecting imputation chains for signs of
non-convergence, but may be challenging to the untrained eye
\citep[\(\S\) 6.5.2]{buur18}. Therefore diagnostic evaluation of
convergence would be preferred. However, not known whether MCMC
diagnostics are appropriate for MICE \ldots{} The application of
convergence diagnostics to MI methods has not been systematically
studied \citep{buur18}. How can non-convergence be diagnosed? How many
iterations are sufficient/required? What are the effects of
non-convergence on estimates, predictions and inference? What is an
appropriate non-convergence diagnostic? Are the default number of
iterations sufficient? Defaults in: SPSS = 10
\href{https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html}{(link)},
Mplus = 100??
\href{https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748}{(link)},
Stata = 10
\href{https://www.stata.com/manuals13/mi.pdf,\%20p.\%20139}{(link)},
Amelia = NA, because resampling, not convergence
\href{https://cran.r-project.org/web/packages/Amelia/Amelia.pdf}{(link)},
en MI = 30
\href{https://cran.r-project.org/web/packages/mi/mi.pdf}{(link)}.

@@

Let \(y\) denote an \(n \times p\) matrix containing the data values on
\(p\) variables for all \(n\) units in a sample. The data value of unit
\(i\) (\(i = 1, \dots, n\)) on variable \(j\) (\(j = 1, \dots, p\)) may
be either observed or missing. The collection of observed data values in
\(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred
to as \(y_{mis}\). \textbf{(This is not only notation, but theory too:
For each datapoint in \(y_{mis}\) \(M \times T\) times plausible values
are sampled, where \(M\) is the number of imputations
(\(m = 1, \dots, M\)) and \(T\) is the number of iterations
(\(t = 1, \dots, T\)).)} The collection of samples between the initial
value (at \(t=1\)) and the imputed value (at \(t=T\)) will be referred
to as an `imputation chain'.

@@

@@

This study evaluates whether \(\widehat{R}\) and \(AC\) could diagnose
convergence of multiple imputation algorithms. We assess the performance
of the two convergence diagnostics against the recommended evaluation
criteria for MI methods \citep[i.e., average bias, average confidence
interval width, and empirical coverage rate across simulations;][\(\S\)
2.5.2]{buur18}. That is, there is no baseline measure available to
evaluate performance against. \textbf{(Move above: The current practice
of visually inspecting imputation chains for signs of non-mixing or
non-stationarity can only diagnose severely pathological cases of
non-convergence.)}

@@

The complete data is `amputed' once for each simulation repetition. That
is, the function \texttt{mice::ampute()} is used to impose a missingness
mechanism upon the data. \textbf{(change this: The missingness is
univariate, and the probability to be missing is the same for all four
variables, namely 20\% (\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This
leaves 20\% of the rows completely observed). }

@@

\textbf{(tot hier doorgelezen)}

@@

where \(t_{(m-1)}\) is the quantile of a \(t\)-distribution with \(m-1\)
degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the
pooled variance estimate. From bias and CIW, we calculate empirical
coverage rates. Coverage rate is the proportion of simulations in which
\(Q\) is between the bounds of the CI95\% around \(\bar{Q}\).

@@

\begin{figure}

{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-2-1}

}

\caption{Convergence diagnostics (means)}\label{fig:unnamed-chunk-2}
\end{figure}

\begin{figure}

{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1}

}

\caption{Convergence diagnostics (variances)}\label{fig:unnamed-chunk-3}
\end{figure}

\begin{figure}

{\centering \includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1}

}

\caption{Convergence diagnostics (regression coefficients)}\label{fig:unnamed-chunk-4}
\end{figure}

\hypertarget{convergence-diagnostics-pair-convergnce-diagnostics-with-respective-estimates-not-split-between-convergence-and-simulation-diagnostics}{%
\subsection{Convergence Diagnostics (pair convergnce diagnostics with
respective estimates, not split between convergence and simulation
diagnostics)}\label{convergence-diagnostics-pair-convergnce-diagnostics-with-respective-estimates-not-split-between-convergence-and-simulation-diagnostics}}

\textbf{(Add different percentages missingness??)}

@@

According to this diagnostic, none of the simulation conditions show
signs of non-convergence, since we only observe negative or zero
\(AC\)-values. The negative \(AC\)-values, however, indicate increasing
divergence (\textbf{move above!}) between subsequent samples in
conditions where \(T<4\). It may not be wise (\textbf{why? and how would
we know? e.g., wider distrubutions? didn't seem like it\ldots{}}) to
terminating the algorithm at that point, but after `recovering' from
this dip in \(AC\)-values. This would imply \(T=6\) as the minimal
number of iterations.

@@

CIW is only clearly divergent in the simulation condition \(T=1\), see
Figure 3B. However, under-estimating the variance of \(\bar{Q}\), which
is the case for \(T=3\), may yield spurious inferences (\textbf{move
above}). Only conditions where \(T>3\) are therefore considered
sufficient. (\textbf{Possibly irrelevant Loess: These conditions have
similar CIWs, but across iterations stability is not established (i.e.,
the Loess line is never completely flat within the 100 simulation
conditions).})

The empirical coverage rate across repetitions seems more or less stable
for conditions where \(T>1\), see Figure 3C. We see some over-coverage
at \(T=2\), but that is better than under-estimating the variance of
\(\bar{Q}\). On average, the coverage rate is somewhat higher than the
expected nominal coverage of 95\% \citep{neym34} (\textbf{not
anymore??}). (\textbf{Loess: Similar to the CIWs, there is some trending
across iterations (i.e., the Loess line is never flat)}).

@@

The observed dip in AC implies that default maxit value of five
iterations is the worst possible number of iterations. Moreover, the
results of this study imply that assessing the stationarity component of
convergence with \(AC\) might be redundant, since high \(AC\)-values are
implausible in MI procedures. That is, the randomness induced by the MI
algorithm effectively mitigates the risk of dependency within chains
(\textbf{add above!!}). \(AC\) would thus not be informative of the
convergence of MI algorithms.

(\textbf{Add above, not here because it was a known limitation: Since
this study only considers only an MCAR missingness mechanism, results
may not be extrapolated to other missing data problems. Proper
performance of the convergence diagnostics under MCAR is necessary but
not sufficient to demonstrate appropriateness of \(\widehat{R}\) and
\(AC\) as convergence diagnostics. This is just a proof of concept.})

>>>>

\begin{abstract}
(\textbf{Rewrite:}) Multple imputation by chained equations (mice) is a
widely used tool to accommodate missing data. While empirical evidence
supports the validity of inferences obtained using mice, there is no
consensus on the convergence properties of the method. This paper
provides insight into non-convergence of mice algorithms.
\end{abstract}

@@

At some point, any social scientist will run into a missing data problem
\citep{alli02}. Missingness is problematic because statistical inference
cannot be performed on incomplete data without employing \emph{ad hoc}
solutions (e.g., list-wise deletion), which may yield wildly invalid
results \citep{buur18}. A popular method to obtain statistically valid
results despite of missing data is `multiple imputation'
\citep[MI;][]{rubin87}. This paper investigates a potential threat to
the validity of this method: algorithmic non-convergence.

With multiple imputation, missing data are `imputed' (i.e., filled in)
several times by means of an iterative algorithmic procedure. The method
ultimately results in an aggregated statistical inference. For the
theoretical foundation of MI, see \citet{rubin87}. For an accessible and
comprehensive introduction to MI from an applied perspective, see
e.g.~\citet{buur18}.

based on a model of the observed databy means of an iterative algorithm
is the use of MICE: multiple imputation by chained equations. MICE is a
type of multiple imputation procedure--a framework proposed by
\citet{rubin87}--in which missing datapoints are `imputed' (i.e., filled
in) by means of an iterative algorithm (chained equations). Imputed
values are obtained by modeling the (distribution of the) missing data
conditional on the observed data. The technique is therefore also called
`fully conditional specification' (FCS).

@@

Current practice is visually inspecting imputation chains for signs of
non-convergence, but may be challenging to the untrained eye
\citep[\(\S\) 6.5.2]{buur18}. Visually inspecting imputation chains
(\textbf{for signs of non-mixing or non-stationarity}) can only diagnose
severely pathological cases of non-convergence. Therefore diagnostic
evaluation of convergence would be preferred. However, not known whether
MCMC diagnostics are appropriate for MICE \ldots{} The application of
convergence diagnostics to MI methods has not been systematically
studied \citep{buur18}. How can non-convergence be diagnosed? How many
iterations are sufficient/required? What are the effects of
non-convergence on estimates, predictions and inference? What is an
appropriate non-convergence diagnostic? Are the default number of
iterations sufficient? Defaults in: SPSS = 10
\href{https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html}{(link)},
Mplus = 100??
\href{https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748}{(link)},
Stata = 10
\href{https://www.stata.com/manuals13/mi.pdf,\%20p.\%20139}{(link)},
Amelia = NA, because resampling, not convergence
\href{https://cran.r-project.org/web/packages/Amelia/Amelia.pdf}{(link)},
en MI = 30
\href{https://cran.r-project.org/web/packages/mi/mi.pdf}{(link)}.

@@

Let \(y\) denote an \(n \times p\) matrix containing the data values on
\(p\) variables for all \(n\) units in a sample. The data value of unit
\(i\) (\(i = 1, \dots, n\)) on variable \(j\) (\(j = 1, \dots, p\)) may
be either observed or missing. The collection of observed data values in
\(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred
to as \(y_{mis}\). (\textbf{This is not only notation, but theory too:}
For each datapoint in \(y_{mis}\) \(M \times T\) times plausible values
are sampled, where \(M\) is the number of imputations
(\(m = 1, \dots, M\)) and \(T\) is the number of iterations
(\(t = 1, \dots, T\)).) The collection of samples between the initial
value (at \(t=1\)) and the imputed value (at \(t=T\)) will be referred
to as an `imputation chain'.

@@

Negative \(AC\)-values indicate divergence between subsequent samples.
It may not be wise (\textbf{why? and how would we know? e.g., wider
distrubutions? didn't seem like it\ldots{}}) to terminating the
algorithm when AC is negative. High \(AC\)-values are implausible in MI
procedures. That is, the randomness induced by the MI algorithm
effectively mitigates the risk of dependency within chains.

@@

This study evaluates whether \(\widehat{R}\) and \(AC\) could diagnose
convergence of multiple imputation algorithms. We assess the performance
of the two convergence diagnostics against the recommended evaluation
criteria for MI methods \citep[i.e., average bias, average confidence
interval width, and empirical coverage rate across simulations;][\(\S\)
2.5.2]{buur18}. That is, there is no baseline measure available to
evaluate performance against.

@@

The complete data is `amputed' once for each simulation repetition. That
is, the function \texttt{mice::ampute()} is used to impose a missingness
mechanism upon the data. (\textbf{change this:} The missingness is
univariate, and the probability to be missing is the same for all four
variables, namely 20\% (\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This
leaves 20\% of the rows completely observed). \textbf{(Add different
percentages missingness??)} This study only considers only an MCAR
missingness mechanism. Therefore, results may not be extrapolated to
other missing data problems. Proper performance of the convergence
diagnostics under MCAR is necessary but not sufficient to demonstrate
appropriateness of \(\widehat{R}\) and \(AC\) as convergence
diagnostics. This is just a proof of concept.

@@

@@

where \(t_{(m-1)}\) is the quantile of a \(t\)-distribution with \(m-1\)
degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the
pooled variance estimate. \textbf{Why track CIW?} Under-estimating the
variance of \(\bar{Q}\) may yield spurious inferences. From bias and
CIW, we calculate empirical coverage rates. Coverage rate is the
proportion of simulations in which \(Q\) is between the bounds of the
CI95\% around \(\bar{Q}\).

@@

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-3-1} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-4-1} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-5-1} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-7-1} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-8-1} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-9-1} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-11-1} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-12-1} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-13-1} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-15-1} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-16-1} \end{flushleft}

\begin{flushleft}\includegraphics{manuscript_files/figure-latex/unnamed-chunk-17-1} \end{flushleft}

\hypertarget{convergence-diagnostics-1}{%
\subsection{Convergence Diagnostics}\label{convergence-diagnostics-1}}

\textbf{(Pair convergnce diagnostics with respective estimates, not
split between convergence and simulation diagnostics)}

@@

According to this diagnostic, none of the simulation conditions show
signs of non-convergence, since we only observe negative or zero
\(AC\)-values.

@@

CIW is only clearly divergent in the simulation condition \(T=1\), see
Figure 3B. Only conditions where \(T>3\) are therefore considered
sufficient. (\textbf{Possibly irrelevant Loess:} These conditions have
similar CIWs, but across iterations stability is not established (i.e.,
the Loess line is never completely flat within the 100 simulation
conditions).)

The empirical coverage rate across repetitions seems more or less stable
for conditions where \(T>1\), see Figure 3C. We see some over-coverage
at \(T=2\), but that is better than under-estimating the variance of
\(\bar{Q}\). On average, the coverage rate is somewhat higher than the
expected nominal coverage of 95\% \citep{neym34} (\textbf{not
anymore??}). (\textbf{Loess:} Similar to the CIWs, there is some
trending across iterations (i.e., the Loess line is never flat)).

@@

The observed dip in AC implies that default maxit value of five
iterations is the worst possible number of iterations. Moreover, the
results of this study imply that assessing the stationarity component of
convergence with \(AC\) might be redundant.
